---
title: "Sea ice extent"
output: html_document
date: "2026-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
```


## Step 1: Import dataset

```{r}
#df <- read.csv("/cloud/project/data/SeaIce_Northern_AllMonths.csv")
#error! This can't be read in beacuse there are some structural issues with the dataset

df <- read.csv("SeaIce_Northern_AllMonths.csv", #file path and file name
               skip = 4, #there are 4 rows above the column names, this is metadata
               header = TRUE) #The first row is column names
```

## Step 2: Clean dataset

```{r}
str(df) #Look at the structure of the new variable df

summary(df) #This function creates a general summary of the dataframe and its columns. Do you see any strange values? 

head(df) #What do the first 10 lines of the data look like? 
```
Let's check the date column. What is the data type of the column "Date"?  
```{r}
str(df$Date)
```

Let's fix it so that it is formatted correctly! We are going to use tidyverse packages to manipulate the dataset and format it so that it is usable. 

In the tidyverse, data are manipulated in sequential steps. Today we will use two functions of the dpylr pacakge: mutate and filter. Mutate creates new columns using other existing columns, and the filter function can help subset data to remove any outliers. 
```{r}
dfClean <- df %>% #start with the original data frame. This %>% is called a pipe, and works to move the object into the next line of code where it can be further manipulated
  mutate(Date = as.Date(Date, format = "%Y%m")) #mutate will help us convert the date column into an actual date object in the correct format 

str(dfClean$Date)

```

Here is a resource for understanding how to format dates in R: https://www.r-bloggers.com/2013/08/date-formats-in-r/